---
title: "Practical Machine Learning Assignment"
author: "Matthew L Maldonado"
date: "April 30, 2016"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Predicting Activity Movement


## Load in libraries that will be needed
```{r}
library(caret)
library(e1071)
```
## Read in the data sets

```{r}
train<-read.csv('pml-training.csv')
test<-read.csv('pml-testing.csv')
```
## Cleaning up the data sets
## Exploring the training set

Let's start by looking at the different columns based off the "str" command.

I will begin by eleiminating the variables with little data to use for predicting the class. I will do this to both the testing and the training set to prepare for the upcoming building and testing of the models.These columns inlucde a lot of blanks and "NA" entries. This is shown with the "str" command below.

```{r}
head(train)
names(train)
str(train,list.len=999)
```
## Remove the Columns that are not needed
```{r}
tidyTrain<-train[,8:length(names(train))]
tidyTest<-test[,8:length(names(test))]


tidyTrain <- subset(tidyTrain, select = -c(max_roll_belt,max_picth_belt,min_roll_belt,min_pitch_belt,amplitude_roll_belt,amplitude_pitch_belt,
                                           var_total_accel_belt,avg_roll_belt,stddev_roll_belt,var_roll_belt,avg_pitch_belt,stddev_pitch_belt,
                                           var_pitch_belt,avg_yaw_belt,stddev_yaw_belt,var_yaw_belt,var_accel_arm,avg_roll_arm,stddev_roll_arm,var_roll_arm,avg_pitch_arm,
                                           stddev_pitch_arm,var_pitch_arm,avg_yaw_arm,stddev_yaw_arm,var_yaw_arm,max_roll_arm,max_picth_arm,max_yaw_arm,min_roll_arm,
                                           min_pitch_arm,min_yaw_arm,amplitude_roll_arm,amplitude_pitch_arm,amplitude_yaw_arm,max_roll_dumbbell,max_picth_dumbbell,
                                           min_roll_dumbbell,min_pitch_dumbbell,amplitude_roll_dumbbell,amplitude_pitch_dumbbell,var_accel_dumbbell,avg_roll_dumbbell,
                                           stddev_roll_dumbbell,var_roll_dumbbell,avg_pitch_dumbbell,stddev_pitch_dumbbell,var_pitch_dumbbell,avg_yaw_dumbbell,
                                           stddev_yaw_dumbbell,var_yaw_dumbbell,max_roll_forearm,max_picth_forearm,min_roll_forearm,min_pitch_forearm,
                                           amplitude_roll_forearm,amplitude_pitch_forearm,var_accel_forearm,avg_roll_forearm,stddev_roll_forearm,var_roll_forearm,
                                           avg_pitch_forearm,stddev_pitch_forearm,var_pitch_forearm,var_pitch_forearm,avg_yaw_forearm,stddev_yaw_forearm,var_yaw_forearm,
                                           kurtosis_roll_belt,kurtosis_picth_belt,kurtosis_yaw_belt,skewness_roll_belt,skewness_roll_belt.1,skewness_yaw_belt,max_yaw_belt,
                                           min_yaw_belt,amplitude_yaw_belt,kurtosis_roll_dumbbell,kurtosis_picth_dumbbell,kurtosis_yaw_dumbbell,skewness_roll_dumbbell,
                                           skewness_pitch_dumbbell,skewness_yaw_dumbbell,max_yaw_dumbbell,min_yaw_dumbbell,amplitude_yaw_dumbbell,kurtosis_roll_forearm,
                                           kurtosis_picth_forearm,kurtosis_yaw_forearm,skewness_roll_forearm,skewness_pitch_forearm,skewness_yaw_forearm,max_yaw_forearm,
                                           min_yaw_forearm,amplitude_yaw_forearm,kurtosis_roll_arm,kurtosis_picth_arm,kurtosis_yaw_arm,skewness_roll_arm,skewness_pitch_arm,
                                           skewness_yaw_arm))

tidyTest <- subset(tidyTest, select = -c(max_roll_belt,max_picth_belt,min_roll_belt,min_pitch_belt,amplitude_roll_belt,amplitude_pitch_belt,
                                           var_total_accel_belt,avg_roll_belt,stddev_roll_belt,var_roll_belt,avg_pitch_belt,stddev_pitch_belt,
                                           var_pitch_belt,avg_yaw_belt,stddev_yaw_belt,var_yaw_belt,var_accel_arm,avg_roll_arm,stddev_roll_arm,var_roll_arm,avg_pitch_arm,
                                           stddev_pitch_arm,var_pitch_arm,avg_yaw_arm,stddev_yaw_arm,var_yaw_arm,max_roll_arm,max_picth_arm,max_yaw_arm,min_roll_arm,
                                           min_pitch_arm,min_yaw_arm,amplitude_roll_arm,amplitude_pitch_arm,amplitude_yaw_arm,max_roll_dumbbell,max_picth_dumbbell,
                                           min_roll_dumbbell,min_pitch_dumbbell,amplitude_roll_dumbbell,amplitude_pitch_dumbbell,var_accel_dumbbell,avg_roll_dumbbell,
                                           stddev_roll_dumbbell,var_roll_dumbbell,avg_pitch_dumbbell,stddev_pitch_dumbbell,var_pitch_dumbbell,avg_yaw_dumbbell,
                                           stddev_yaw_dumbbell,var_yaw_dumbbell,max_roll_forearm,max_picth_forearm,min_roll_forearm,min_pitch_forearm,
                                           amplitude_roll_forearm,amplitude_pitch_forearm,var_accel_forearm,avg_roll_forearm,stddev_roll_forearm,var_roll_forearm,
                                           avg_pitch_forearm,stddev_pitch_forearm,var_pitch_forearm,var_pitch_forearm,avg_yaw_forearm,stddev_yaw_forearm,var_yaw_forearm,
                                           kurtosis_roll_belt,kurtosis_picth_belt,kurtosis_yaw_belt,skewness_roll_belt,skewness_roll_belt.1,skewness_yaw_belt,max_yaw_belt,
                                           min_yaw_belt,amplitude_yaw_belt,kurtosis_roll_dumbbell,kurtosis_picth_dumbbell,kurtosis_yaw_dumbbell,skewness_roll_dumbbell,
                                           skewness_pitch_dumbbell,skewness_yaw_dumbbell,max_yaw_dumbbell,min_yaw_dumbbell,amplitude_yaw_dumbbell,kurtosis_roll_forearm,
                                           kurtosis_picth_forearm,kurtosis_yaw_forearm,skewness_roll_forearm,skewness_pitch_forearm,skewness_yaw_forearm,max_yaw_forearm,
                                           min_yaw_forearm,amplitude_yaw_forearm,kurtosis_roll_arm,kurtosis_picth_arm,kurtosis_yaw_arm,skewness_roll_arm,skewness_pitch_arm,
                                           skewness_yaw_arm))

```


## Create training and test sets

Next I will break up the testing set even further and create a new training and testing set using cross validation.This is bassically 1-fold cross validation. The smaller K should produce more bias. The previous testing set is then held out for after the model is built and validated.

```{r}
set.seed(353)
inTrain <- createDataPartition(y=tidyTrain$classe, p = 0.7, list = FALSE)
training <- tidyTrain[inTrain,]
testing <- tidyTrain[-inTrain,]
```
 We'll first train a Decision Tree and an SVM.
```{r,cache=TRUE}
## Training a Model
modFitDT <- train(classe~., data = training, method ="rpart")
predDT <- predict(modFitDT,testing)
cmDT <- confusionMatrix(predDT, testing$classe)
cmDT


modelSVM <- svm(classe~ ., data = training)
predSVM <- predict(modelSVM,testing)
cmSVM <- confusionMatrix(predSVM, testing$classe)
cmSVM
```
The decsion tree has rather poor performance so we'll move on to using an SVM for our final test set.

Next we'll predict on our validation set which we will submit for grading. I estimate the out of sample error will be in the 90-95% range in line with our testing set.
```{r}
predict(modelSVM,tidyTest)

```
This would also agree with the result of the quiz which is 18/20 correct.




